---
title: "chapter 5"
author: "Arto Vesterbacka"
date: "25 11 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(dplyr)
library(GGally)
library(corrplot)
library(ggplot2)
library(FactoMineR)
```

### Data Wrangling

```{r data wrangling}
hd <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human_development.csv", stringsAsFactors = F)
gii <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/gender_inequality.csv", stringsAsFactors = F, na.strings = "..")
str(gii)
str(hd)
summary(gii)
summary(hd)

hd<- hd %>% rename(
  hd.index = Human.Development.Index..HDI.,
  hd.life = Life.Expectancy.at.Birth,
  hd.education.exp = Expected.Years.of.Education,
  hd.education.mean = Mean.Years.of.Education,
  hd.GDP =  Gross.National.Income..GNI..per.Capita,
  hd.gni.rank = GNI.per.Capita.Rank.Minus.HDI.Rank
  )
gii<-gii %>% rename(
  gii.index = Gender.Inequality.Index..GII.,
  gii.mom.death = Maternal.Mortality.Ratio,
  gii.ad.births = Adolescent.Birth.Rate,
  gii.reps = Percent.Representation.in.Parliament,
  gii.2ndedu.f = Population.with.Secondary.Education..Female.,
  gii.2ndedu.m = Population.with.Secondary.Education..Male.,
  gii.labor.f = Labour.Force.Participation.Rate..Female.,
  gii.labor.m = Labour.Force.Participation.Rate..Male.
  )
summary(gii)
gii <- mutate(gii, edu.ratio = gii.2ndedu.f/gii.2ndedu.m)
gii <- mutate(gii, lab.ratio = gii.labor.f/gii.labor.m)

humandata <- inner_join(hd, gii, by = "Country")
str(humandata)

humandata$hd.GDP <- as.numeric(gsub(",", ".", gsub("\\.", "", as.character(humandata$hd.GDP))))
keep_columns = c("Country", "gii.2ndedu.f", "gii.labor.f", "hd.education.exp", "hd.life", "hd.GDP", "gii.mom.death", "gii.ad.births", "gii.reps")

human <- select(humandata, one_of(keep_columns))
human_ <- filter(human, complete.cases(human))
last <- nrow(human_) - 7

# choose everything until the last 7 observations
human_ <- human_[0:last, ]

# add countries as rownames
rownames(human_) <- human_$Country
human_ <- select(human_, -Country)
str(human_)
summary(human_)
```

## Principal Component Analysis

```{r analysis PCA}

cor(human_) %>% corrplot
pca_human <- prcomp(human_)
summary(pca_human)
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
human_std <- scale(human_)
pca_human_std <- prcomp(human_std)
summary(pca_human_std)
biplot(pca_human_std, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```
The data shows high positive correlation between life expectancy at birth and 2nd degree education, also for expected education. A negative correlation is between mother's death at childbirth and 2nd degree education. Mother's death at childbirth is a negative correlation for life expectancy at birth and expected education.

Scaling the variables reduces the spread of the data. This also makes the output harder to interprit. Without scaling there can be a lot of variability, and higher numerical values tend to skew the output more. 


## Multiple Correspondence Analysis

```{r analysis MCA}

data(tea)
# column names to keep in the dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- select(tea, one_of(keep_columns))
# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA

plot(mca, invisible=c("ind"), habillage = "quali")


```

First dimension is heavily influenced by unpackageed and tea shop. These are also fairly high on the second dimension but negatively. Second dimension is positively effected by other and chain store+tea shop.