---
title: "chapter 3"
output: html_document
---

```{r setupp, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Logistic regression

Arto Vesterbacka

This chapter is about logistic regression using student data.

## Data Wrangling

Here we take the student data from two files, combine them into one with only the observations that are found from
both sets. We also create new variables from existing ones to be used in later analysis.

### Setting up libraries and working directory.

Here we load all the required libraries for the data wrangling and analysis.

```{r Libraries}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)
```

### Setting the working directory

Here we set the working directory. If you are going to run this on your computer, you will need to set tis
to the correct path corresponding to your system.

```{r Working directory}
setwd("C:/koulu/mooc/intro to open data/IODS-project/") # change this if you wish to run the code on your computer
```

### Loading the data

Loading the data from 2 csv-files, and checking their dimensions and structures

```{r Data Loading}
dta_mat <- read.csv("student-mat.csv", sep=";", header=TRUE)
dta_por <- read.csv("student-por.csv", sep=";", header=TRUE)
dim(dta_mat)
dim(dta_por)
str(dta_mat)
str(dta_por)
```

### Joining the two dataset

Joining the two dataset with only the students present in both sets

```{r Joining the datasets}
join_by <- c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet")
mat <- inner_join(dta_mat, dta_por, by = join_by)
dim(mat)
str(mat)
```

### Data concatination

Here we find all the variables from the 2 data sets and combine them by taking the average of the numerical
variables, and taking only one of the non-numerical variables

```{r if_else_structure}

alc <- select(mat, one_of(join_by))

# the columns in the datasets which were not used for joining the data
notjoined_columns <- colnames(dta_mat)[!colnames(dta_mat) %in% join_by]

# for every column name not used for joining...
for(column_name in notjoined_columns) {
  # select two columns from 'math_por' with the same original name
  two_columns <- select(mat, starts_with(column_name, ignore.case = TRUE))
  # select the first column vector of those two columns
  first_column <- select(two_columns, 1)[[1]]

  # if that first column vector is numeric...
  if(is.numeric(first_column)) {
    # take a rounded average of each row of the two columns and
    # add the resulting vector to the alc data frame
    alc[column_name] <- round(rowMeans(two_columns))
  } else { # else if it's not numeric...
    # add the first column vector to the alc data frame
    alc[column_name] <- first_column
  }
}

```
### Mutating columns into one

Here we create new variables from existing ones to be used in the later stages of the analysis

```{r mutations}
alc <- mutate(alc, alc_use = (Dalc + Walc) / 2)
alc <- mutate(alc, high_use = alc_use > 2)
glimpse(alc)
```
### Save the data

Here we save the data into a csv-file for later usage.

```{r saving data to csv}
write.csv(x = alc, file="./data/alc_data.csv")
```

## Analysis

### Reading the data from the csv-file

```{r read data}
chap3 <- read.table("./data/alc_data.csv", sep=",", header=TRUE)
str(chap3)
```

### Selecting 4 interesting variable for analysis on alcohol usage

We select absences, romantic, goout, and internet as the variables to explain high/low alcohol usage

```{r EDA}
chap3 %>% group_by(sex, high_use, romantic, internet) %>% summarise(count = n(), absens = mean(absences, go_out = mean(goout)))
par(mfrow=c(2,2))
pair_plot <- ggpairs(chap3, columns = c("sex","absences", "romantic", "goout", "internet"), aes(col = sex, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
pair_plot

```

### Logistic regression model

Here we will do the logistic regression analysis using our selected variables

```{r logistic regression}
model = glm(high_use ~ absences + romantic + goout + internet, data = chap3, family = "binomial")
summary(model)
OR = coef(model) %>% exp
CI = confint(model) %>% exp
cbind(OR, CI)
```
